\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=2.5cm]{geometry}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{titlesec}

\titleformat{\section}{\normalfont\Large\bfseries}{}{0pt}{}
\titleformat{\subsection}{\normalfont\large\bfseries}{}{0pt}{}

\title{\textbf{Project Contribution Report}}
\author{Elnaz Azizi, Charly Watts, Christian Unterrainer}
\date{\today}

\begin{document}

\maketitle

\section*{Team Contributions Overview}

The following outlines the individual contributions of each team member to the project. All members participated actively and collaboratively in meetings and feedback discussions.

\section*{Elnaz Azizi (\texttt{GitHub: Elnaazi, Matrnr: 3434538})}

\begin{itemize}[leftmargin=1.5em]
    \item Organized initial project meetings.
    \item Implemented machine learning classifiers: Naive Bayes, Random Forest, and XGBoost.
    \item Researched and evaluated additional classification models, with a focus on optimizing performance using XGBoost.
    \item Applied data imputation strategies to effectively handle missing values.
    \item Focused on performance optimization by integrating F1-threshold tuning to improve classification quality
    \item Incorporated improvements based on professor's feedback (notably imputation techniques).
    \item Responsible for writing the \textit{Methods} section of the paper (implementation description).
\end{itemize}

\section*{Charly Watts (\texttt{GitHub: BytebenderCoco, Matrnr: 3422867})}

\begin{itemize}[leftmargin=1.5em]
    \item Initiated the project and coordinated progress and team meetings.
    \item Created and managed the GitHub repository and uploaded relevant data files.
    \item Implemented machine learning classifiers: k-Nearest Neighbors (k-NN) and Multi-Layer Perceptron (MLP).
    \item Refactored and structured the main notebook for improved code readability and visualization.
    \item Adjusted dataset splitting based on feedback.
    \item Interpreted professor feedback and defined corresponding tasks.
    \item Conducted hyperparameter tuning for model optimization.
    \item Resolved technical issues (e.g., XGBoost warnings).
    \item Standardized the F2-score computation within the notebook.
    \item Organized the structure and formatting of the paper (language and template).
    \item Contributed content to the paper, including \textit{Related Work}, tables, and citations.
    \item Authored the remaining sections of the paper, including the \textit{Introduction}, \textit{Conclusion}, and others.
\end{itemize}

\section*{Christian Unterrainer (\texttt{GitHub: cunterrainer, Matrnr: 3430428})}

\begin{itemize}[leftmargin=1.5em]
    \item Proposed the original project idea with a focus on Health/Heart Disease.
    \item Implemented code components and supported experimental workflows.
    \item Merged baseline classifiers into a single notebook and streamlined the process by adding pipelines.
    \item Implemented machine learning classifiers: Logistic regression and the support vector machine (SVM)
    \item Carried out experiments based on professor feedback, including:
    \begin{itemize}
        \item Implemented emphasizing Recall over the F1-score.
        \item Comparing model performance by tuning parameters e.g. tree counts for random forest or XGBoost.
        \item Investigating the impact of columns with significant missing values by trying to remove them and merging all dataset into a larger single dataset.
    \end{itemize}
    \item Reported and documented the results of all experiments.
    \item Responsible for writing the \textit{Experiments} section of the paper (experiment setup, results, discussion) and the contribution paper.
\end{itemize}

\end{document}
